---
title: "Compute per-barcode polyspecificity reactivity"
author: "Tyler Starr"
date: "8/9/2021"
output:
  github_document:
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

This notebook reads in per-barcode counts from `count_variants.ipynb` for PSR-binding mini-Tite-seq experiments, computes functional scores for binding of each barcoded variant to the PSR reagent, and does some basic QC on variant binding functional scores.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))
options(repos = c(CRAN = "https://cran.r-project.org"))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","fitdistrplus")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#make output directory
if(!file.exists(config$PSR_bind_dir)){
  dir.create(file.path(config$PSR_bind_dir))
}
```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

## Setup
First, we will read in metadata on our sort samples, the table giving number of reads of each barcode in each of the sort bins, and the barcode-variant lookup tables, and merge these tables together.

```{r input_data}
#read dataframe with list of barcode runs
barcode_runs <- read.csv(file=config$barcode_runs,stringsAsFactors=F); barcode_runs <- subset(barcode_runs, select=-c(R1))

#eliminate rows from barcode_runs that are not from a PSR binding experiment
barcode_runs <- barcode_runs[barcode_runs$sample_type == "PSR",]

#read file giving count of each barcode in each sort partition
counts <- data.table(read.csv(file=config$variant_counts_file,stringsAsFactors=F))

#eliminate rows from counts that are not part of an titration bin sample
counts <- subset(counts, sample %in% barcode_runs[barcode_runs$sample_type=="PSR","sample"])

#read in barcode-variant lookup tables
dt <- data.table(read.csv(file=config$codon_variant_table_file,stringsAsFactors=F))

dt <- merge(counts, dt, by=c("library","barcode")) ;rm(counts)
```

 Convert from Illumina read counts to estimates of the number of cells that were sorted into a bin, and add some other useful information to our data frame.
 
```{r downweight_counts_by_cells}
#for each bin, normalize the read counts to the observed ratio of cell recovery among bins
for(i in 1:nrow(barcode_runs)){
  lib <- as.character(barcode_runs$library[i])
  bin <- as.character(barcode_runs$sample[i])
  ratio <- sum(dt[library==lib & sample==bin,"count"])/barcode_runs$number_cells[i]
  if(ratio<1){ #if there are fewer reads from a FACS bin than cells sorted
    dt[library==lib & sample==bin, count.norm := as.numeric(count)] #don't normalize cell counts, make count.norm the same as count
    print(paste("reads < cells for",lib,bin,", un-normalized (ratio",ratio,")")) #print to console to inform of undersampled bins
  }else{
    dt[library==lib & sample==bin, count.norm := as.numeric(count/ratio)] #normalize read counts by the average read:cell ratio, report in new "count.norm" column
    print(paste("read:cell ratio for",lib,bin,"is",ratio))
  }
}

#annotate each barcode as to whether it's a homolog variant, SARS-CoV-2 wildtype, synonymous muts only, stop, nonsynonymous, >1 nonsynonymous mutations
dt[,variant_class:=as.character(NA)]
dt[n_codon_substitutions==0, variant_class := "wildtype"]
dt[n_codon_substitutions > 0 & n_aa_substitutions==0, variant_class := "synonymous"]
dt[n_aa_substitutions>0 & grepl("*",aa_substitutions,fixed=T), variant_class := "stop"]
dt[n_aa_substitutions == 1 & !grepl("*",aa_substitutions,fixed=T), variant_class := "1 nonsynonymous"]
dt[n_aa_substitutions > 1 & !grepl("*",aa_substitutions,fixed=T), variant_class := ">1 nonsynonymous"]

#cast the data frame into wide format
dt <- dcast(dt, library + barcode + target + variant_class + aa_substitutions + n_aa_substitutions ~ sample, value.var="count.norm")

```

## Calculating mean bin for each barcode at each sample concentration
Next, for each barcode at each of the PSR concentrations, calculate the "mean bin" response variable. We will use a maximum likelihood approach to determine the mean and standard deviation of fluorescence for a barcode, given its distribution of cell counts across sort bins, and the known fluorescence boundaries of those sort bins from the sorting log. The package `fitdistcens` enables this ML estimation for these type of *censored* observations, where we know we observed a cell within some fluorescence interval but do not know the exact fluorescence value attributed to that observation. The counts are multiplied by 20 so that there is not a large rounding effect when they are rounded to integers.

Fluorescence boundaries of the sort bins

```
(-288, 183), (184, 892), (893, 3892), (3893, 262143)
```

```{r calculate_mean_bin, error=FALSE, message=FALSE, warning=FALSE, results=F}
#add total count corresponding to count across the four bins for each barcode. 
dt[,psr_count_01 := sum(PSR_01_bin1,PSR_01_bin2,PSR_01_bin3,PSR_01_bin4),by=c("library","barcode")]
#add indicator if count>1 in >1 bin
dt[,total_bins_w_count_01 := sum(.(PSR_01_bin1,PSR_01_bin2,PSR_01_bin3,PSR_01_bin4)>0),by=c("library","barcode")]

#add total count corresponding to count across the four bins for each barcode. 
dt[,psr_count_02 := sum(PSR_02_bin1,PSR_02_bin2,PSR_02_bin3,PSR_02_bin4),by=c("library","barcode")]
#add indicator if count>1 in >1 bin
dt[,total_bins_w_count_02 := sum(.(PSR_02_bin1,PSR_02_bin2,PSR_02_bin3,PSR_02_bin4)>0),by=c("library","barcode")]

#add total count corresponding to count across the four bins for each barcode. 
dt[,psr_count_03 := sum(PSR_03_bin1,PSR_03_bin2,PSR_03_bin3,PSR_03_bin4),by=c("library","barcode")]
#add indicator if count>1 in >1 bin
dt[,total_bins_w_count_03 := sum(.(PSR_03_bin1,PSR_03_bin2,PSR_03_bin3,PSR_03_bin4)>0),by=c("library","barcode")]

#add total count corresponding to count across the four bins for each barcode. 
dt[,psr_count_04 := sum(PSR_04_bin1,PSR_04_bin2,PSR_04_bin3,PSR_04_bin4),by=c("library","barcode")]
#add indicator if count>1 in >1 bin
dt[,total_bins_w_count_04 := sum(.(PSR_04_bin1,PSR_04_bin2,PSR_04_bin3,PSR_04_bin4)>0),by=c("library","barcode")]

#add total count corresponding to count across the four bins for each barcode. 
dt[,psr_count_05 := sum(PSR_05_bin1,PSR_05_bin2,PSR_05_bin3,PSR_05_bin4),by=c("library","barcode")]
#add indicator if count>1 in >1 bin
dt[,total_bins_w_count_05 := sum(.(PSR_05_bin1,PSR_05_bin2,PSR_05_bin3,PSR_05_bin4)>0),by=c("library","barcode")]



#define function to calculate ML meanF
calc.MLmean <- function(b1,b2,b3,b4,min.b1,min.b2,min.b3,min.b4,max.b4,min.count=1){ #b1-4 gives observed cell counts in bins 1-4; remaining arguments give fluorescence boundaries of the respective bins; min.count gives minimum number of total observations needed across bins in order to calculate meanF (default 1)
  data <- data.frame(left=c(rep(min.b1,round(b1)),rep(min.b2,round(b2)),rep(min.b3,round(b3)),rep(min.b4,round(b4))),
                     right=c(rep(min.b2,round(b1)),rep(min.b3,round(b2)),rep(min.b4,round(b3)),rep(max.b4,round(b4)))) #define data input in format required for fitdistcens
  if(nrow(unique(data))>1 & nrow(data)>min.count){ #only fits if above user-specified min.count, and if the data satisfies the fitdistcens requirement that cells are observed in at least two of the censored partitions to enable ML estimation of identifiable parameters
    fit <- fitdistcens(data,"norm")
    return(list(as.numeric(summary(fit)$estimate["mean"]),as.numeric(summary(fit)$estimate["sd"])))
  } else {
    return(list(as.numeric(NA),as.numeric(NA)))
  }
}

#fit ML mean and sd fluorescence for each barcode, and calculate total cell count as the sum across the four bins. Multiply cell counts by a factor of 20 to minimize rounding errors since fitdistcens requires rounding to integer inputs
invisible(dt[,c("polyspecificity_01","PSR_sdF_01") := tryCatch(calc.MLmean(b1=PSR_01_bin1*20,b2=PSR_01_bin2*20,
                                                                      b3=PSR_01_bin3*20,b4=PSR_01_bin4*20,
                                                                      min.b1=log(20),min.b2=log(183.5),min.b3=log(892.5),
                                                                      min.b4=log(3892),max.b4=log(229000)),
                                                          error=function(e){return(list(as.numeric(NA),as.numeric(NA)))}),by=c("library","barcode")])

invisible(dt[,c("polyspecificity_02","PSR_sdF_02") := tryCatch(calc.MLmean(b1=PSR_02_bin1*20,b2=PSR_02_bin2*20,
                                                                      b3=PSR_02_bin3*20,b4=PSR_02_bin4*20,
                                                                      min.b1=log(20),min.b2=log(183.5),min.b3=log(892.5),
                                                                      min.b4=log(3892),max.b4=log(229000)),
                                                          error=function(e){return(list(as.numeric(NA),as.numeric(NA)))}),by=c("library","barcode")])

invisible(dt[,c("polyspecificity_03","PSR_sdF_03") := tryCatch(calc.MLmean(b1=PSR_03_bin1*20,b2=PSR_03_bin2*20,
                                                                      b3=PSR_03_bin3*20,b4=PSR_03_bin4*20,
                                                                      min.b1=log(20),min.b2=log(183.5),min.b3=log(892.5),
                                                                      min.b4=log(3892),max.b4=log(229000)),
                                                          error=function(e){return(list(as.numeric(NA),as.numeric(NA)))}),by=c("library","barcode")])

invisible(dt[,c("polyspecificity_04","PSR_sdF_04") := tryCatch(calc.MLmean(b1=PSR_04_bin1*20,b2=PSR_04_bin2*20,
                                                                      b3=PSR_04_bin3*20,b4=PSR_04_bin4*20,
                                                                      min.b1=log(20),min.b2=log(183.5),min.b3=log(892.5),
                                                                      min.b4=log(3892),max.b4=log(229000)),
                                                          error=function(e){return(list(as.numeric(NA),as.numeric(NA)))}),by=c("library","barcode")])

invisible(dt[,c("polyspecificity_05","PSR_sdF_05") := tryCatch(calc.MLmean(b1=PSR_05_bin1*20,b2=PSR_05_bin2*20,
                                                                      b3=PSR_05_bin3*20,b4=PSR_05_bin4*20,
                                                                      min.b1=log(20),min.b2=log(183.5),min.b3=log(892.5),
                                                                      min.b4=log(3892),max.b4=log(229000)),
                                                          error=function(e){return(list(as.numeric(NA),as.numeric(NA)))}),by=c("library","barcode")])


#save temp data file for downstream troubleshooting since the ML meanF took >1hr to calculate -- don't use these for final anlaysis though for reproducibility!
save(dt,file=paste(config$PSR_bind_dir,"/dt.temp.Rda",sep=""))
#load(file=paste(config$PSR_bind_dir,"/dt.temp.Rda",sep=""))
```

## Basic plotting and QC

Some stop variants eked through our scFv+ selection gate, remove the remaining.
```{r remove_stops}
dt[variant_class == "stop",c("polyspecificity_01","polyspecificity_02","polyspecificity_03","polyspecificity_04","polyspecificity_05") := list(as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]

dt[variant_class == "stop",c("psr_count_01","psr_count_02","psr_count_03","psr_count_04","psr_count_05") := list(as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA),as.numeric(NA))]
```

Let's look at the distibution of expression scores by variant class for each library.

```{r unfiltered_polyspecificity_distribution, echo=T, fig.width=5, fig.height=15, fig.align="center", dpi=300,dev="png"}
par(mfrow=c(5,1))
#histogram of mean psr bin, separated by class
hist(dt[variant_class %in% (c("1 nonsynonymous",">1 nonsynonymous")),polyspecificity_01],col="gray40",main="",breaks=50,xlab="PSR labeling (1x)")
hist(dt[variant_class %in% (c("synonymous","wildtype")),polyspecificity_01],col="#92278F",add=T,breaks=50)

hist(dt[variant_class %in% (c("1 nonsynonymous",">1 nonsynonymous")),polyspecificity_02],col="gray40",main="",breaks=50,xlab="PSR labeling (0.316x)")
hist(dt[variant_class %in% (c("synonymous","wildtype")),polyspecificity_02],col="#92278F",add=T,breaks=50)

hist(dt[variant_class %in% (c("1 nonsynonymous",">1 nonsynonymous")),polyspecificity_03],col="gray40",main="",breaks=50,xlab="PSR labeling (0.1x)")
hist(dt[variant_class %in% (c("synonymous","wildtype")),polyspecificity_03],col="#92278F",add=T,breaks=50)

hist(dt[variant_class %in% (c("1 nonsynonymous",">1 nonsynonymous")),polyspecificity_04],col="gray40",main="",breaks=50,xlab="PSR labeling (0.0316x)")
hist(dt[variant_class %in% (c("synonymous","wildtype")),polyspecificity_04],col="#92278F",add=T,breaks=50)

hist(dt[variant_class %in% (c("1 nonsynonymous",">1 nonsynonymous")),polyspecificity_05],col="gray40",main="",breaks=50,xlab="PSR labeling (0x)")
hist(dt[variant_class %in% (c("synonymous","wildtype")),polyspecificity_05],col="#92278F",add=T,breaks=50)
```


Next let's look at the distributon of cell counts across the four bins for each barcode. 
```{r cell_count_coverage, echo=T, fig.width=5, fig.height=15, fig.align="center", dpi=300,dev="png"}
#histograms
par(mfrow=c(5,1))
hist(log10(dt[,psr_count_01]+0.1),xlab="",main="1x",col="gray50")
hist(log10(dt[,psr_count_02]+0.1),xlab="",main="0.316x",col="gray50")
hist(log10(dt[,psr_count_03]+0.1),xlab="",main="0.1x",col="gray50")
hist(log10(dt[,psr_count_04]+0.1),xlab="",main="0.0316x",col="gray50")
hist(log10(dt[,psr_count_05]+0.1),xlab="cell count (log10, plus 0.1 pc)",main="0x",col="gray50")

```
Filter out expression measurements determined from <10 estimated cells

```{r filter_min_cfu}
min_count <- 10
dt[psr_count_01<min_count, c("polyspecificity_01","psr_count_01") := NA]
dt[psr_count_02<min_count, c("polyspecificity_02","psr_count_02") := NA]
dt[psr_count_03<min_count, c("polyspecificity_03","psr_count_03") := NA]
dt[psr_count_04<min_count, c("polyspecificity_04","psr_count_04") := NA]
dt[psr_count_05<min_count, c("polyspecificity_05","psr_count_05") := NA]

```

Look at correlations in PSR labeling for each of the five concentrations of PSR reagent.

```{r psr_conc_correlations, echo=T, fig.width=15, fig.height=15, fig.align="center", dpi=300,dev="png"}
pairs(dt[,.(polyspecificity_01,polyspecificity_02,polyspecificity_03,polyspecificity_04,polyspecificity_05)], main="",pch=19,col="#00000010")

invisible(dev.print(pdf, paste(config$PSR_bind_dir,"/correlations_PSR_concentrations.pdf",sep=""),useDingbats=F))

```

Look at PSR 02 as violin plots, faceted by each target. In next notebook, we'll evaluate count depth and possibly apply further filtering to remove low-count expression estimates

```{r psr_distribution_vioplot, echo=T, fig.width=6, fig.height=4.5, fig.align="center", dpi=300,dev="png"}
p1 <- ggplot(dt[!is.na(polyspecificity_02),],aes(x=variant_class,y=polyspecificity_02))+
  geom_violin(scale="width")+stat_summary(fun=median,geom="point",size=1)+
  ggtitle("polyspecificity (0.316x PSR)")+xlab("variant class")+theme(axis.text.x=element_text(angle=-90,hjust=0))+
  facet_wrap(~target,nrow=4)

grid.arrange(p1,ncol=1)

#save pdf
invisible(dev.print(pdf, paste(config$PSR_bind_dir,"/violin-plot_PSR-by-target.pdf",sep="")))
```


We have generated PSR binding measurements for `r round(nrow(dt[!is.na(polyspecificity_02)])/nrow(dt)*100,digits=2)`% of the barcodes in our libraries. (0.316x PSR)

## Data Output

Finally, let's output our measurements for downstream analyses.

```{r output_data}
dt[,.(library,barcode,target,variant_class,aa_substitutions,n_aa_substitutions,
     psr_count_01,polyspecificity_01,
     psr_count_02,polyspecificity_02,
     psr_count_03,polyspecificity_03,
     psr_count_04,polyspecificity_04,
     psr_count_05,polyspecificity_05)] %>%
  mutate_if(is.numeric, round, digits=6) %>%
  write.csv(file=config$PSR_bind_file, row.names=F)

```



