---
title: "Map DMS to the chIgY-Fab sturcture"
author: "Tyler Starr"
date: "4/29/2022"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---
This notebook analyzes the structure of the CGGnaive Fab bound to the chicken IgY dimer. It generates a list of inter-residue distances and maps sitewise DMS properties to the B-factor column for visualization in PyMol.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","gridExtra","bio3d","ggrepel")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#make output directory
if(!file.exists(config$structural_mapping_dir)){
  dir.create(file.path(config$structural_mapping_dir))
}

```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

## Data input

Read in tables of mutant measurements.

```{r input_data}
dt <- data.table(read.csv(file=config$final_variant_scores_mut_file,stringsAsFactors=F))
```

Read in the structure pdb

```{r input_pdb}
sites <- read.csv(file=config$CGGnaive_site_info,stringsAsFactors=F)
```

Read in the table giving annotations and information on CGGnaive sites
```{r input_sites}
pdb <- read.pdb(file=config$pdb)
```


## Compute site-wise metrics from DMS data for mapping to structure

Want to map the mean effect of mutations on each of the three properties to structure. Also plot the *max* effect of any mutation at a site on binding as a metric of evolvability. Only include mutations with n_bc > 3 for these properties.
```{r compute_site_properties}
sites$mean_CGG_bind <- NA
sites$max_CGG_bind <- NA
sites$quartile75_CGG_bind <- NA
sites$mean_expression <- NA
sites$mean_polyspecificity <- NA
  
for(i in 1:nrow(sites)){
  if(sites[i,"chain"] != "link"){
    sites$mean_CGG_bind[i] <- mean(dt[position_IMGT==sites[i,"site"] & chain==sites[i,"chain"] & wildtype != mutant & n_bc_bind_CGG>3,delta_bind_CGG],na.rm=T)
    sites$max_CGG_bind[i] <- max(dt[position_IMGT==sites[i,"site"] & chain==sites[i,"chain"] & wildtype != mutant & n_bc_bind_CGG>3,delta_bind_CGG],na.rm=T)
    sites$quartile75_CGG_bind[i] <- quantile(dt[position_IMGT==sites[i,"site"] & chain==sites[i,"chain"] & wildtype != mutant & n_bc_bind_CGG>3,delta_bind_CGG],0.75,na.rm=T)
    sites$mean_expression[i] <- mean(dt[position_IMGT==sites[i,"site"] & chain==sites[i,"chain"] & wildtype != mutant & n_bc_expr>3,delta_expr],na.rm=T)
    sites$mean_polyspecificity[i] <- mean(dt[position_IMGT==sites[i,"site"] & chain==sites[i,"chain"] & wildtype != mutant & n_bc_psr>3,delta_psr],na.rm=T)
  }
}

```


## Map metrics to the structure
```{r map_properties_to_pdb}
b_mean_CGG <- rep(0, length(pdb$atom$b))
b_max_CGG <- rep(0, length(pdb$atom$b))
b_75th_CGG <- rep(0, length(pdb$atom$b))
b_mean_expr <- rep(0, length(pdb$atom$b))
b_mean_psr <- rep(0, length(pdb$atom$b))
for(i in 1:nrow(pdb$atom)){
  res <- pdb$atom$resno[i]
  if(pdb$atom$chain[i] %in% c("B","H") & res %in% sites[sites$chain=="H","site"]){
    b_mean_CGG[i] <- sites[sites$site==res & sites$chain=="H","mean_CGG_bind"]
    b_max_CGG[i] <- sites[sites$site==res & sites$chain=="H","max_CGG_bind"]
    b_75th_CGG[i] <- sites[sites$site==res & sites$chain=="H","quartile75_CGG_bind"]
    b_mean_expr[i] <- sites[sites$site==res & sites$chain=="H","mean_expression"]
    b_mean_psr[i] <- sites[sites$site==res & sites$chain=="H","mean_polyspecificity"]
  }else if(pdb$atom$chain[i] %in% c("C","L") & res %in% sites[sites$chain=="L","site"]){
    b_mean_CGG[i] <- sites[sites$site==res & sites$chain=="L","mean_CGG_bind"]
    b_max_CGG[i] <- sites[sites$site==res & sites$chain=="L","max_CGG_bind"]
    b_75th_CGG[i] <- sites[sites$site==res & sites$chain=="L","quartile75_CGG_bind"]
    b_mean_expr[i] <- sites[sites$site==res & sites$chain=="L","mean_expression"]
    b_mean_psr[i] <- sites[sites$site==res & sites$chain=="L","mean_polyspecificity"]
  }
}
write.pdb(pdb=pdb, file=paste(config$structural_mapping_dir,"/CGG_mean.pdb",sep=""), b=b_mean_CGG)
write.pdb(pdb=pdb, file=paste(config$structural_mapping_dir,"/CGG_max.pdb",sep=""), b=b_max_CGG)
write.pdb(pdb=pdb, file=paste(config$structural_mapping_dir,"/CGG_75th.pdb",sep=""), b=b_75th_CGG)
write.pdb(pdb=pdb, file=paste(config$structural_mapping_dir,"/expression_mean.pdb",sep=""), b=b_mean_expr)
write.pdb(pdb=pdb, file=paste(config$structural_mapping_dir,"/psr_mean.pdb",sep=""), b=b_mean_psr)

```

Generate a list of contact residues (5A or closer contacts)

Below are heavy chain residues that are 5A or closer to CGG ligand in each of the two protomers:

```{r heavy_contacts}
binding.site(pdb,
             a.inds=atom.select(pdb,chain=c("H","B")),
             b.inds=atom.select(pdb,chain=c("A","D")),
             cutoff=5,hydrogens=F)$resnames

```

Below are light chain residues that are 5A or closer to CGG ligand in each of the two protomers:

```{r light_contacts}
binding.site(pdb,
             a.inds=atom.select(pdb,chain=c("L","C")),
             b.inds=atom.select(pdb,chain=c("A","D")),
             cutoff=5,hydrogens=F)$resnames

```

Below are heavy chain residues that are 5A or closer to light chain in each of the two protomers

```{r heavy_contacts_with_light}
binding.site(pdb,
             a.inds=atom.select(pdb,chain=c("H","B")),
             b.inds=atom.select(pdb,chain=c("L","C")),
             cutoff=5,hydrogens=F)$resnames

```

Below are light chain residues that are 5A or closer to heavy chain in each of the two protomers

```{r light_contacts_with_heavy}
binding.site(pdb,
             a.inds=atom.select(pdb,chain=c("L","C")),
             b.inds=atom.select(pdb,chain=c("H","B")),
             cutoff=5,hydrogens=F)$resnames

```